---
title: "How Humans See Data"
author: "Marieke Jones"
date: "3/5/2020"
output: html_document
---

This repository hosts code creating all the plots in my presentation How Humans See Data, originally presented as a Learning Shot for iTHRIV at the University of Virginia.

The code these plots are based on is from John Rauser's How Humans See Data, presented at [Velocity Amsterdam 2016](http://conferences.oreilly.com/velocity/devops-web-performance-eu/public/schedule/detail/54354). Original code on github at [https://github.com/jrauser/writing/tree/master/how_humans_see_data](https://github.com/jrauser/writing/tree/master/how_humans_see_data).

```{r}
library(tidyverse)
library(viridis)
library(scales)
library(ggrepel)
```

# Part 1: Why visualize at all?

## A cat

Showing these data as a table 

```{r}
cat <- read_csv("cat.csv")
cat
```

Make a boring picture of i against a and b. Does not show the intended relationship between A and B

```{r}
cat %>%
  pivot_longer(cols = c(a,b), names_to = "var", values_to = "value") %>%
  ggplot() + 
  geom_line(aes(i, value, col = var), size = 2) +
  theme_bw() +
  scale_color_manual(values = c("midnightblue", "chocolate2")) +
  labs(color = "")

ggsave("plots/01_cat_lines.png", width = 5, height = 3.5)
```

Make the cat. Showing these data in a scatterplot makes the relationship obivous pre-attentively.

```{r}
ggplot(cat, aes(a,b)) + 
  geom_point(col = "midnightblue", size=2) + 
  coord_equal() +
  theme_classic()

ggsave("plots/02_cat.png", width = 5, height = 3.5)
```

# Color

Really made up of 3 channels

```{r}
values <- data.frame(val = 1:5, y = 1)

ggplot(values, aes(val,y,fill=val)) + 
  geom_tile() + 
  scale_fill_gradient(low="black", high="white", limits=c(0,6)) +
  theme_classic() +
  theme(axis.title = element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(),
        legend.position="none")

ggsave("plots/02_01_luminance.png", height = 3.5/3, width = 5)

ggplot(values, aes(val,y,fill=val)) + 
  geom_tile() + 
  scale_fill_gradient(low="white", high="red", limits=c(0,6)) +
  theme_classic() +
  theme(axis.title = element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(),
        legend.position="none")

ggsave("plots/02_02_saturation.png", height = 3.5/3, width = 5)

ggplot(values, aes(val,y,fill=factor(val))) + 
  geom_tile() + 
  scale_fill_discrete() + 
  theme_classic() +
  theme(axis.title = element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(),
        legend.position="none")

ggsave("plots/02_03_color.png", height = 3.5/3, width = 5)

```


## Hue

Hue encodes TB incidence per 100k
TB dataset showing African countries by average tb incidence per 100K -- Alphabetical sort

```{r}
tb <- read_csv("tb.csv")

# shorten two longest country names
tb <- tb %>%
  mutate(country = forcats::fct_recode(country, 
                          Tanzania = "United Republic of Tanzania",
                          DRC = "Democratic Republic of the Congo"))

# create summary table that we'll use for many plots
AFR_sum <- tb %>%
  filter(who_region == "AFR") %>%
  group_by(country) %>%
  summarize(TBper100k = mean(incidence_100k, na.rm = TRUE))

# using hue without a natural order
AFR_sum %>%
  ggplot(aes("a", country, fill = TBper100k)) + 
  geom_tile() + 
  scale_fill_gradient2(midpoint=median(AFR_sum$TBper100k), mid="#ffffbf", low="#fc8d59", high="#91bfdb") +
  theme_classic() +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_text(size=6),
        axis.ticks.x=element_blank())
ggsave("plots/03_tb_hue1.png", width = 5, height = 3.5)

# using viridis hue scale which has intuitive order due to added channel of color luminance
 AFR_sum %>%
  ggplot(aes("a", country, fill = TBper100k)) + 
  geom_tile() + 
  scale_fill_viridis() + 
  theme_classic() +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_text(size=6),
        axis.ticks.x=element_blank())
ggsave("plots/03_tb_hue2.png", width = 5, height = 3.5)
```

Reorderd intellegently

```{r}
AFR_sum %>%
  ggplot(aes("a", reorder(country, TBper100k), fill = TBper100k)) + 
  geom_tile() + 
  scale_fill_gradient2(midpoint=median(AFR_sum$TBper100k), mid="#ffffbf", low="#fc8d59", high="#91bfdb") +
  theme_classic() +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_text(size=6),
        axis.ticks.x=element_blank())

ggsave("plots/04_tb_hue1_ordered.png", width = 5, height = 3.5)

AFR_sum %>%
  ggplot(aes("a", reorder(country, TBper100k), fill = TBper100k)) + 
  geom_tile() + 
  scale_fill_viridis() + 
  theme_classic() +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_text(size=6),
        axis.ticks.x=element_blank())
ggsave("plots/04_tb_hue2_ordered.png", width = 5, height = 3.5)
```

## Saturation of one color

Ordered and with default "low" value

```{r}
AFR_sum %>%
  ggplot(aes("a", reorder(country, TBper100k), fill = TBper100k)) + 
  geom_tile() + 
  scale_fill_gradient(low="white", high=muted("blue")) +
  theme_classic() +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_text(size=6),
        axis.ticks.x=element_blank())

ggsave("plots/05_tb_saturation.png", width = 5, height = 3.5)
```

Now with 0 as white
For accurate ratioing with saturation, the scale has to extend to zero.

```{r}
AFR_sum %>%
  ggplot(aes("a", reorder(country, TBper100k), fill = TBper100k)) + 
  geom_tile() + 
  scale_fill_gradient(low="white", high=muted("blue")) +
  expand_limits(fill = 0) +
  theme_classic() +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_text(size=6),
        axis.ticks.x=element_blank())

ggsave("plots/06_tb_saturation_zero.png", width = 5, height = 3.5)


# what is the incidence in Mauritius?
AFR_sum %>%
  arrange(TBper100k)
# 12.8
```

## Area

```{r}
#remove two countries to make an even grid of 45 countries
tb_ordered <- tb %>%
  filter(who_region == "AFR") %>%
  filter(country != "Angola" & country != "Senegal") %>%  
  group_by(country) %>%
  summarize(TBper100k = mean(incidence_100k, na.rm = TRUE)) %>%
  mutate(country = reorder(country, TBper100k)) %>%
  arrange(TBper100k, country)

tb_ordered

tb_ordered %>%
  ggplot(aes(rep(1:5, 9), rep(1:9, each = 5))) +
  geom_point(aes(size = TBper100k)) +
  geom_text(aes(label = country), position = position_nudge(y=0.5), size=2) +
  expand_limits(x=c(0.5,5.5), y=c(1,8.5)) + 
  ylab(NULL) + 
  scale_size_area() + #with zero as zero incidence
  theme_classic() +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
ggsave("plots/07_tb_area.png", width = 5, height = 3.5)

```

Accurate ratioing depends on a scale that goes to zero, so making sure that that is the case for this plot is important

## Angle

Again, note that the scale goes to zero (a horizontal line).

```{r}
make_segment <- function(df, scale_low, scale_high) {
  scale_range <- scale_high-scale_low
  fraction <- (df$TBper100k-scale_low) / scale_range
  theta <- fraction * pi/2
  return(data.frame(x=c(0, cos(theta)),
                    y=c(0, sin(theta))))
}

tb_angle <- AFR_sum %>%
  filter(country != "Angola" & country != "Senegal") %>%
  group_by(country) %>%
  do(make_segment(., 0, 1057.333)) %>%
  ungroup()

ggplot(tb_angle, aes(x, y, group=country))+
  geom_path() +
  facet_wrap(~country, ncol = 9) + coord_equal()

# reorder country and clean up the plot
tb_angle %>%
  mutate(country = factor(country, levels = unique(tb_ordered$country))) %>%
  ggplot(aes(x, y, group=country))+
  geom_path() +
  facet_wrap(~country, ncol = 9) + 
  coord_equal() +
  theme_bw() +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        strip.text.x = element_text(size = 5))

ggsave("plots/08_tb_angle.png", width = 5, height = 3.5)
```

Angle plot that we look at all the time
```{r}
tb %>%
  group_by(who_region, year) %>%
  summarise(population = mean(pop)/1e6) %>%
  ggplot(aes(year, population, color = who_region)) +
  geom_smooth(method = "lm", se = FALSE) + 
  scale_y_log10(labels = comma) +
  theme_bw() +
  labs(x = NULL, y = "Population (Millions)", color = "Region")

ggsave("plots/08_tb_pop_angle.png", width = 5, height = 3.5)
```

## Length

Length not on a common scale. Add a bit of random jitter to the position of the beginning of the bar

```{r}
# not ordered
AFR_sum$random <- runif(n = 47, min = 0, max = 100)

AFR_sum %>%
  ggplot(aes(x=random, xend=TBper100k + random, y = country, yend = country)) +
  geom_segment(size = 2) + xlab(NULL) + ylab(NULL) + theme_bw() + theme(axis.text.y=element_text(size=6))

ggsave("plots/09_tb_length.png", width = 5, height = 3.5)

# ordered
tb_ordered$random <- runif(n = 45, min = 0, max = 100)

tb_ordered %>%
  ggplot(aes(x=random, xend=TBper100k + random, y = country, yend = country)) +
  geom_segment(size = 2) + 
  theme_bw() + 
  theme(axis.text.y=element_text(size=6),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

ggsave("plots/09_tb_length.png", width = 5, height = 3.5)
```

Looking at length not on common baseline is actually quite common

```{r}
countries <- c("Togo", "Rwanda", "Sao Tome and Principe", "Eritrea", "Niger", "Equatorial Guinea", "Mauritania", "Burundi", "Ghana", "Guinea")

AFR_error <- tb %>%
  filter(who_region == "AFR") %>%
  filter(country %in% countries) %>%
  group_by(country) %>%
  summarize(TBper100k = mean(incidence_100k, na.rm = TRUE),
            SD_TB = sd(incidence_100k, na.rm = TRUE)) %>%
  mutate(country = reorder(country, TBper100k)) %>%
  arrange(TBper100k, country)

ggplot(AFR_error, aes(x = country, y = TBper100k)) +
  geom_pointrange(aes(ymin = TBper100k - SD_TB, 
                      ymax = TBper100k + SD_TB)) +
  theme_bw() +
  coord_flip() +
  theme(axis.title.y=element_blank())
  
ggsave("plots/10_tb_length_error.png", width = 5, height = 3.5)
```

## Position on misaligned scales

```{r}
tb_ordered %>%
  head(25) %>%
  ggplot(aes(x = TBper100k, y="a")) +
  geom_point(size = 2) + 
  scale_x_continuous() +
  facet_wrap(~country, ncol=5) +
  theme_bw() + 
  theme(strip.text.x = element_text(size = 8),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_text(size = 6),
        axis.ticks.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.major.x = element_line(colour = "grey80", size = 0.4),
        panel.grid.minor.x = element_line(colour = "grey80", size = 0.4, linetype="dotted"))

ggsave("plots/11_tb_position_facet.png", width = 5, height = 3.5)

```

Time series faceted on same but misaligned scale

```{r}
tb %>%
  filter(who_region == "WPR") %>%
  filter(country== "Niue" | country == "Tuvalu") %>%
  ggplot(aes(year, incidence_100k)) + 
  geom_line(aes(group = country, color = country), size = 1.5) +
  facet_wrap(~country) +
  theme_bw() + 
  labs(y = "TB Incidence per 100K") +
  theme(strip.text.x = element_text(size = 10),
        axis.text.x = element_text(size = 8),
        axis.title.x = element_blank(),
        legend.position = "none")

ggsave("plots/12_tb_time_facet.png", width = 5, height = 3.5)
```

## Position on common scale

```{r}
ggplot(tb_ordered, aes(x = TBper100k, y = country)) + 
  geom_point() +
  theme_bw() + 
  labs(x = "TB Incidence per 100K") +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_text(size = 6))

ggsave("plots/13_tb_point.png", width = 5, height = 3.5)

# unordered
AFR_sum %>%
  filter(country != "Angola" & country != "Senegal") %>%
  ggplot(aes(x = TBper100k, y = country)) + 
  geom_point() +
  theme_bw() + 
  labs(x = "TB Incidence per 100K") +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_text(size = 6))

ggsave("plots/14_tb_point_unordered.png", width = 5, height = 3.5)
```

Should we include zero?

On the tb plot we did because the range of incidence was roughly 12 - 1200. In a different dataset, zero is not necessary
```{r}
mtcars %>%
  mutate(model = reorder(model, mpg)) %>%
  arrange(mpg, model) %>%
  ggplot(aes(x=mpg, y=model)) + 
  geom_point() + 
  ylab(NULL) + 
  expand_limits(x=0) +
  theme_bw() + 
  labs(x = "Miles per gallon") +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_text(size = 6))

ggsave("plots/15_cars_point_zero.png", width = 5, height = 3.5)
```

# Other Observations

## Stacked anything is nearly always a mistake

A stacked bar chart does two things, neither of them well.

```{r}
ggplot(diamonds, aes(clarity, fill=cut, group=cut)) + 
  geom_bar(stat="count", position="stack") +
  scale_y_continuous("Count", labels=comma) +
  theme_classic()

ggsave("plots/16_diamonds_stacked.png", width = 5, height = 3.5)
```

Instead, let's see the same plot with parallel lines

```{r}
diamonds_summary <- diamonds %>%
  group_by(clarity, cut) %>%
  summarize(n = n())

ggplot(diamonds_summary, aes(clarity, n, color=cut, group=cut)) + 
  geom_line(lwd = 1.25) + 
  geom_point(size = 2) + 
  scale_y_continuous("Count", labels=comma) +
  theme_classic()

ggsave("plots/17_diamonds_parallel.png", width = 5, height = 3.5)

# total count per clarity/cut class
diamonds_tot <- diamonds %>%
  group_by(clarity) %>%
  summarize(n=n())

ggplot() + 
  geom_line(data = diamonds_summary, aes(clarity, n, color=cut, group=cut), lwd = 1.25) + 
  geom_point(data = diamonds_summary, aes(clarity, n, color=cut, group=cut), size = 2) + 
  geom_line(data = diamonds_tot, aes(clarity, n), group=1, lwd = 1.25, color = "gray50") + 
  geom_point(data = diamonds_tot, aes(clarity, n), size = 2, color = "gray50") +
  scale_y_continuous("Count", labels=comma) +
  theme_classic()

ggsave("plots/18_diamonds_total.png", width = 5, height = 3.5)

```

# Part 3: Assembly

## The Law of Similarity

Color is pretty good

```{r}
# Make number of cylinders a factor, so ggplot will choose a discrete scale.
mtcars$cyl <- factor(mtcars$cyl)

ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(color=cyl), size = 2.5) + 
  xlab("weight (1,000 lbs)") +
  ylab("Miles per gallon") +
  scale_color_viridis_d("Cylinders") +
  theme_classic()

ggsave("plots/19_cars_color.png", width = 5, height = 3.5)

```

Shape can be ok, but choose wisely

```{r}
# closed shapes
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(shape=cyl), size = 2.5) + 
  xlab("weight (1,000 lbs)") +
  ylab("Miles per gallon") +
  scale_shape("Cylinders") +
  theme_classic()

ggsave("plots/20_cars_shape.png", width = 5, height = 3.5)

# open shapes
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(shape=cyl), size = 2.5) + 
  xlab("weight (1,000 lbs)") +
  ylab("Miles per gallon") +
  scale_shape_manual("Cylinders", values = c(3,2,1)) +
  theme_classic()

ggsave("plots/21_cars_shape.png", width = 5, height = 3.5)
```

You'd think this would be a great encoding, but it's terrible because of the similar curvature of the sixes and eights.

```{r}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(shape=cyl), size = 2.5) + 
  xlab("weight (1,000 lbs)") +
  ylab("Miles per gallon") +
  scale_shape_manual("Cylinders", values = c(52,54,56)) +
  theme_classic()

ggsave("plots/21A_cars_shape2.png", width = 5, height = 3.5)
```

Encode information in 2 channels: hue and shape

```{r}
# open shapes
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(shape=cyl, color = cyl), size = 2.5) + 
  xlab("weight (1,000 lbs)") +
  ylab("Miles per gallon") +
  scale_shape_manual("Cylinders", values = c(3,2,1)) +
  scale_color_viridis_d("Cylinders") +
  theme_classic()

ggsave("plots/22_cars_hue_shape.png", width = 5, height = 3.5)
```

Overlaying linear fits encloses space and makes the grouped objects very strong.

```{r}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(shape=cyl, color=cyl), size = 2.5) + 
  geom_smooth(aes(group=cyl, color=cyl), method="lm") + 
  xlab("weight (1,000 lbs)") +
  ylab("Miles per gallon") +
  scale_shape_manual("Cylinders", values = c(3,2,1)) +
  scale_color_viridis_d("Cylinders") +
  theme_classic()

ggsave("plots/23_cars_hue_shape.png", width = 5, height = 3.5)
```

Law of proximity

```{r}
ggplot(diamonds, aes(clarity, fill=cut, group=cut)) + 
  geom_bar(stat="count", position="dodge") +
  scale_y_continuous("Count", labels=comma) +
  theme_classic()

ggsave("plots/24_diamonds_grouped.png", width = 5, height = 3.5)
```

# Detection

Each point is a car
```{r}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(size = 2.5) + 
  xlab("weight (1,000 lbs)") +
  ylab("Miles per gallon") +
  theme_classic()

ggsave("plots/25_detection.png", width = 5, height = 3.5)
```

Each point is a country year
```{r}
set.seed(403)
tb %>%
  sample_n(300) %>%
  ggplot(aes(pop/1e6, incidence_100k)) + 
  geom_point() + 
  xlab("Population (Millions)") +
  ylab("TB incidence per 100k") +
  scale_x_log10(labels = comma) +
  theme_bw()

ggsave("plots/26_TB_detection.png", width = 5, height = 3.5)
```

Too little luminance in the points

```{r}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(size = 2.5, color = "gray90") + 
  xlab("weight (1,000 lbs)") +
  ylab("Miles per gallon") +
  theme_classic()

ggsave("plots/27_detection.png", width = 5, height = 3.5)
```

Too small points

```{r}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(size = .5) + 
  xlab("weight (1,000 lbs)") +
  ylab("Miles per gallon") +
  theme_classic()

ggsave("plots/28_detection.png", width = 5, height = 3.5)
```

Gridlines too thick

```{r}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point() + 
  xlab("weight (1,000 lbs)") +
  ylab("Miles per gallon") +
  theme_bw() +
  theme(panel.grid.minor.y = element_line(color = "gray20", size = 1),
        panel.grid.major.y = element_line(color = "gray20", size = 2),
        panel.grid.major.x = element_line(color = "gray20", size = 2),
        panel.grid.minor.x = element_line(color = "gray20", size = 1))

ggsave("plots/29_detection.png", width = 5, height = 3.5)
```

Gridlines
```{r}
tb %>%
  filter(country== "Marshall Islands" | country == "Tuvalu" | country == "Kiribati" | country == "Palau") %>%
  mutate(country = factor(country, levels = c("Kiribati", "Marshall Islands", "Tuvalu", "Palau"))) %>%
  ggplot(aes(year, incidence_100k)) + 
  geom_line(aes(group = country, color = country), size = 1.5) +
  facet_wrap(~country) +
  theme_classic() + 
  labs(y = "TB Incidence per 100K") +
  theme(strip.text.x = element_text(size = 10),
        axis.text.x = element_text(size = 8),
        axis.title.x = element_blank(),
        legend.position = "none")

ggsave("plots/30_tb_time_facet.png", width = 5, height = 3.5)

tb %>%
  filter(country== "Marshall Islands" | country == "Tuvalu" | country == "Kiribati" | country == "Palau") %>%
  mutate(country = factor(country, levels = c("Kiribati", "Marshall Islands", "Tuvalu", "Palau"))) %>%
  ggplot(aes(year, incidence_100k)) + 
  geom_line(aes(group = country, color = country), size = 1.5) +
  facet_wrap(~country) +
  theme_bw() + 
  labs(y = "TB Incidence per 100K") +
  theme(strip.text.x = element_text(size = 10),
        axis.text.x = element_text(size = 8),
        axis.title.x = element_blank(),
        legend.position = "none")

ggsave("plots/31_tb_time_facet.png", width = 5, height = 3.5)

# same plot without facets
tb %>%
  filter(country== "Marshall Islands" | country == "Tuvalu" | country == "Kiribati" | country == "Palau") %>%
  mutate(country = factor(country, levels = c("Kiribati", "Marshall Islands", "Tuvalu", "Palau"))) %>%
  ggplot(aes(year, incidence_100k)) + 
  geom_line(aes(group = country, color = country), size = 1.5) +
  #facet_wrap(~country) +
  theme_bw() + 
  labs(y = "TB Incidence per 100K") +
  theme(strip.text.x = element_text(size = 10),
        axis.text.x = element_text(size = 8),
        axis.title.x = element_blank())

ggsave("plots/32_tb_time.png", width = 5, height = 3.5)
```

Include 0 -- yes with length of bars

```{r}
tb_ordered %>%
  ggplot(aes(x=TBper100k, y = country)) +
  geom_col(fill = "black") + 
  theme_bw() +
  labs(x = "TB incidence per 100k") +
  theme(axis.text.y=element_text(size=6),
        axis.title.y=element_blank())

ggsave("plots/33_tb_length.png", width = 5, height = 3.5)

# now with cars
mtcars %>%
  mutate(model = reorder(model, mpg)) %>%
  arrange(mpg, model) %>%
  ggplot(aes(x=mpg, y=model)) + 
  geom_col(fill = "black") + 
  theme_bw() +
  labs(x = "Miles per gallon") +
  theme(axis.text.y=element_text(size=6),
        axis.title.y=element_blank())

ggsave("plots/34_cars_length.png", width = 5, height = 3.5)
```

